### 2013-05-02 doc/dissertation/document/backmatter/pn_equations.tex
adding pn and spn appendices

%        File: spn.tex
\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}
\usepackage{listings}
\usepackage{textcomp}
\definecolor{listinggray}{gray}{0.9}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{
  backgroundcolor=\color{lbcolor},
  tabsize=4,
  rulecolor=,
  language=c++,
  basicstyle=\scriptsize,
  upquote=true,
  aboveskip={1.5\baselineskip},
  columns=fixed,
  showstringspaces=false,
  extendedchars=true,
  breaklines=true,
  prebreak =
  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
  frame=single,
  showtabs=false,
  showspaces=false,
  showstringspaces=false,
  identifierstyle=\ttfamily,
  keywordstyle=\color[rgb]{0,0,1},
  commentstyle=\color[rgb]{0.133,0.545,0.133},
  stringstyle=\color[rgb]{0.627,0.126,0.941},
}

\author{Stuart R. Slattery
  \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

\date{\today}
\title{A Monte Carlo Synthetic Acceleration Method for the $SP_N$ Equations}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\section{Introduction}
\label{sec:introduction}
The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \cite{evans_denovo:_2010}. For full reactor
core neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the repsonses of interest. Other forms of the
transport problem, including the $P_N$ method, take on a simpler form
than the more common $S_N$ methods but lack in accuracy when compared
while still requiring considerable computational resources for
solutions in multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
methods was either widely unknown, widely unused, or combination of
both even though numerical studies showed promising results with
better solutions than diffusion theory and a significant reduction in
computational time over more accurate methods such as discrete
ordinates. Why did this happen? A significant problem, pointed out by
Larsen, was that little rigor had been applied to the formulation of
the $SP_N$ equations since their derivation through primarily
heuristic arguments. Instead, studies at that time focused on simply
comparing the results of the method to other contemporary transport
solution strategies. In addition, many problems of interest from the
literature at the time were either solved using nodal-type methods for
reactor-sized problems or $S_N$-type methods for benchmark problems
with intricate material configurations and potentially large flux
gradients over small spatial domains.

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment with both variational and
asymptotic derivations performed as a means of verification. In
addition, these equations have been more rigorously studied as
solution methods to MOX fuel problems and have been shown to provide
accurate solutions. With this mathematical literature to provide a
solid numerical footing for the method, we look at its application to
today's challenge problems in full reactor core transport. The
reduction in numerical complexity of current full core deterministic
solution methods using the $S_N$ approximation could mean significant
savings in both compute time and memory required. In addition, the
characteristics of the solution to the transport problem for a steady
state reactor core permit diffusion theory to be used; a staple of the
nuclear industry since its inception. Therefore, if diffusion theory
is applicable, then finer grained solutions that capture more of the
physics contained in the transport equation should be possible with
the $SP_N$ method. In doing so, we also expect from the literature to
obtain computed responses on the order of accuracy we would expect
from an appropriately discretized $S_N$ method at a fraction of the
cost.

To further motivate moving in this direction, recent developments in
the Exnihilo neutronics package at Oak Ridge National Laboratory have
permitted generation of the $SP_N$ system of equations for detailed
full core reactor models. By fully forming these equations and
formulating them as a linear algebra problem, we now have access to
all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and preconditioning
methods such as algebraic multigrid. This leads us to then explore the
applicability of our work in discrete Monte Carlo methods for linear
systems as a possible solution method for the $SP_N$ equations. If
formulated correctly, we hypothesize that signficant improvement in
usage of computational resources may be observed compared to modern
solution techniques such as those suggested. In addition, solving the
$SP_N$ equations in this way also breaks away from the $S_N$ forms of
parallelism where spatial parallelism is achieved by an efficient
parallel sweep, angular efficiency achieved by pipelining, and energy
parallelism achieved by decoupling the groups. With the $SP_N$
equations as a full matrix system, we now can parallelize the problem
as prescribed by the linear solver, which may be significantly more
scalable than current $S_N$ transport practices.

In this document we derive the multigroup $SP_N$ equations from the
Boltzmann transport equation. We then perform a spectral analysis on
the resulting system of equations to determine its potential
performance with both modern Krylov methods and Monte Carlo methods
for asymmetric linear systems. Following this, we devise a Monte Carlo
synthetic acceleration method as a proposed solution scheme.

%%---------------------------------------------------------------------------%%
\section{Derivation of the $SP_N$ Equations}
\label{sec:derivation}
In this section, we derive the $SP_N$ equations, closely following the
work of Evans. We begin by stating the general time-independent
neutron transport equation followed by a derivation of the $P_N$
equations in planar geometry for multiple energy groups. From these
equations, we then apply a set of approximations to yield the
multi-dimensional, multi-group $SP_N$ equations for fixed source
problems.

\subsection{The Neutron Transport Equation}
\label{subsec:transport_eq}
As a starting point we define the time-independent neutron transport
equation \cite{lewis_computational_1993}:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \int \int
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \rightarrow
  \hat{\Omega}) \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
  q(\vec{r},\hat{\Omega},E)\:,
  \label{eq:general_transport}
\end{multline}
with the variables defined as:
\begin{itemize}
\item $\vec{r}$ - neutron spatial position
\item $\hat{\Omega}$ - neutron streaming direction with radial
  component $\mu$ and azimuthal component $\omega$
\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
  scattering
\item $E$ - neutron energy
\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
\item $\sigma(\vec{r},E)$ - total interaction cross section
\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
  of scattering from direction $\hat{\Omega}'$ into an angular domain
  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
  $E'$ to an energy domain $dE'$ about energy $E$
\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
\end{itemize}
For this work, it is sufficient to formulate
Eq~(\ref{eq:general_transport}) in 1-dimensonal Cartesian geometry:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \int \int \sigma_s(x,E') \rightarrow
  E,\hat{\Omega}' \rightarrow \hat{\Omega}) \psi(x,\hat{\Omega}',E')
  d\Omega' dE' + \frac{q(x,E)}{4 \pi}\:,
  \label{eq:cart_1d_transport}
\end{multline}
where the angular component of the solution is no longer dependent on
the azimuthal direction of travel and an isotropic source of neutrons
is assumed.

\subsection{Derivation of the $P_N$ Equations}
\label{subsec:pn_equations}
Next, we derive the $P_N$ equations, a simplified form of the general
transport equation where Legendre polynomials are used to expand the
angular flux and scattering cross section variables as a means of
capturing the angular structure of the solution. Before deriving this
form of the transport equation, we briefly discuss a few properties of
Legendre polynomials that we will find useful in the derivation.

\subsubsection{Legendre Polynomials}
\label{subsubsec:legendre_polys}
The Legendre polynomials are an orthogonal set of functions that
are solutions to Legendre's differential equation. They have the
following form \cite{lewis_computational_1993}:
\begin{equation}
  P_l(\mu) = \frac{1}{2^l l!}\frac{d^l}{d \mu^l}(\mu^2-1)^l\:.
  \label{eq:general_legendre_poly}
\end{equation}
These functions have several useful properties including
orthogonality:
\begin{equation}
  \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = \frac{1}{2l+1}\delta_{l l'}\:,
  \label{eq:legendre_orthog}
\end{equation}
a recurrence relation:
\begin{equation}
  \mu P_l(\mu) = \frac{1}{2l+1}[(l+1)P_{l+1}(\mu) + l P_{l-1}(\mu)]\:,
  \label{eq:legendre_recurrence}
\end{equation}
and an addition theorem:
\begin{equation}
  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}\sum_{m=-l}^l
  Y_{lm}(\hat{\Omega})Y^*_{lm}(\hat{\Omega}')\:,
  \label{eq:legendre_addition}
\end{equation}
where the functions $Y_{lm}(\hat{\Omega})$ are the spherical
harmonics. We can form the addition theorem in this way because the
spherical harmonics are in fact just harmonic multiples of the
Legendre polynomials:
\begin{equation}
  Y_{lm}(\hat{\Omega}) =
  \sqrt{\frac{(2l+1)(l-m)!}{(l+m)!}}P^m_l(\mu)e^{i \omega m}\:,
  \label{eq:spherical_harmonic}
\end{equation}
where $\omega$ is the azimuthal component of the streaming
direction. We can reduce Eq~(\ref{eq:legendre_addition}) for the
planar geometry we are studying by ignoring the azimuthal components
of the addition theorem. As shown in Eq~(\ref{eq:spherical_harmonic}),
the azimuthal dependence is given by the harmonic component, $e^{i
  \omega m}$, and therefore we choose to ignore all terms in
Eq.~(\ref{eq:legendre_addition}) where $m \neq 0$. This gives:
\begin{equation}
  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}
  Y_{l0}(\hat{\Omega})Y^*_{l0}(\hat{\Omega}')\:.
  \label{eq:legendre_addition_2}
\end{equation}
Per Eq~(\ref{eq:spherical_harmonic}) we have:
\begin{equation}
  Y_{l0}(\hat{\Omega}) = \sqrt{2l+1}P_l^0(\mu)\:,
  \label{eq:harmonic_0}
\end{equation}
where $P_l^0(\mu) = P_l(\mu)$ is the $0^{th}$ associated Legendre
function. Finally, we can reduce the addition theorem from
Eq~(\ref{eq:legendre_addition_2}) with Eq~(\ref{eq:harmonic_0}) to a
simple product for planar geometry:
\begin{equation}
  P_l(\hat{\Omega} \cdot \hat{\Omega}') = P_l(\mu)P_l(\mu')\:.
  \label{eq:legendre_addition_3}
\end{equation}

\subsubsection{Planar $P_N$ Equations}
\label{subsubsec:planar_pn}
With the Legendre polynomial properties defined above, we can proceed
by deriving the $P_N$ equations for planar geometry. For these
equations, we will start by assuming a monoenergtic field of neutrons
such that we are solving the following reduced form of the transport
equation: 
\begin{equation}
  \mu \frac{\partial}{\partial x} \psi(x,\mu) + \sigma(x) \psi(x,\mu)
  = \int d\Omega' \sigma_s(x,\hat{\Omega}' \rightarrow \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}') + \frac{q(x)}{4 \pi}\:.
  \label{eq:mono_transport}
\end{equation}
The $P_N$ equations introduce the approximation that the angular
dependence of the scattering cross section, $\sigma_s$, and the angular
flux $\psi$, can be discretized by expanding them in Legendre
polynomials as follows:
\begin{equation}
  \psi(x,\mu) = \sum_{n=0}^\infty (2n+1)P_n(\mu)\phi_n(x)\:,
  \label{eq:flux_expansion}
\end{equation}
\begin{equation}
  \sigma_{sm}(x) = \sum_{m=0}^\infty (2m+1)P_m(\mu)\sigma_s(x)\:,
  \label{eq:scattering_expansion}
\end{equation}
where we have supressed the $2\pi$ generated by integrating away the
azimuthal angular component and $\phi_n(x)$ in
Eq~(\ref{eq:flux_expansion}) is referred to as the $n^{th}$ Legendre
moment of the neutron flux and is given by:
\begin{equation}
  \phi_n(x) = \int_{-1}^1 P_n(\mu)\psi(x,\mu)d\mu\:.
  \label{eq:legendre_moments}
\end{equation}
We first insert the expansions given by Eq~(\ref{eq:flux_expansion})
and (\ref{eq:scattering_expansion}) into the planar transport equation
given by Eq~(\ref{eq:mono_transport}):
\begin{multline}
  \frac{\partial}{\partial x}\Big[\sum_{n=0}^\infty (2n+1) \phi_n \mu
    P_n(\mu) \Big] + \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) =
  \\ \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d \mu' + q\:,
  \label{eq:pn_deriv_1}
\end{multline}
where the dependence on the spatial variable, $x$, has been
supressed. To arrive at the $P_N$ equations we multiply
Eq~(\ref{eq:pn_deriv_1}) by $P_m(\mu)$ and integrate over the angular
domain $\int_{-1}^1 d \mu$. We will look at each term in
Eq~(\ref{eq:pn_deriv_1}) individually.

@@

Now that we have expanded all angular dependent terms in
Eq~(\ref{eq:pn_deriv_1}) and reduced them appropriately, we can
combine them to generate the $P_N$ equations:
\begin{equation}
    \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{2n+1} \Big[
      (n+1) \phi_{n+1} + n \phi_{n-1} \Big] + \sum_{n=0}^\infty \sigma
    \phi_n = \sum_{n=0}^\infty \sigma_{sn} \phi_n + q\delta_{n0}\:.
  \label{eq:pn_deriv_18}
\end{equation}
More formally, the $P_N$ equations are written as:
\begin{equation}
   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
  \label{eq:final_pn_equations}
\end{equation}
where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
yields a set of $N+1$ equations for $N+2$ flux moments. We therefore
require an additional equation to close the system. In accordance with
the series truncation as an approximation we choose the last moment in
the expansion to be zero:
\begin{equation}
  \phi_{N+1} = 0\:.
  \label{eq:pn_closure}
\end{equation}
As an example, we will construct the P5 equations from
Eq~(\ref{eq:final_pn_equations}) and the closure given by
Eq~(\ref{eq:pn_closure}):
\begin{subequations}
  \begin{gather}
   \frac{\partial}{\partial x}\phi_{1} + \Sigma_0 \phi_0 = q\:,\\ 
   \frac{1}{3} \frac{\partial}{\partial x}\Big[ 2
     \phi_{2} + \phi_{0} \Big] + \Sigma_1 \phi_1 = 0\:,\\
   \frac{1}{5} \frac{\partial}{\partial x}\Big[ 3 \phi_{3} + 2
     \phi_{1} \Big] + \Sigma_2 \phi_2 = 0 \:,\\
   \frac{1}{7} \frac{\partial}{\partial x}\Big[ 4 \phi_{4} + 3
     \phi_{2} \Big] + \Sigma_3 \phi_3 = 0 \:,\\
   \frac{1}{9} \frac{\partial}{\partial x}\Big[ 5 \phi_{5} + 4
     \phi_{3} \Big] + \Sigma_4 \phi_4 = 0 \:,\\
   \frac{1}{11} \frac{\partial}{\partial x} 5 \phi_{4} + \Sigma_5
   \phi_5 = 0 \:.
  \end{gather}
  \label{eq:p5_equations}
\end{subequations}
This gives us a set of 6 coupled equations for the six Legendre
moments requested defined over the entire spatial domain for a single
energy group. In practice, only odd-numbered $P_N$ orders are
generally used \cite{lewis_computational_1993}. This is due to the
fact that using odd $N$ yields an even number of $N+1$ equations which
can be split evenly on the left and right boundaries of the problem to
facilitate the description of boundary conditions. We will choose this
convention when deriving the boundary conditions.

\subsubsection{Boundary Conditions for the $P_N$ Equations}
\label{subsbusec:bcs_pn}
Per the analysis in \cite{lewis_computational_1993}, two types of
boundary conditions will be discussed: reflecting and Marshak. Marshak
boundary conditions can be used to specify both vacuum conditions and
isotropic source conditions on the boundary.

@@

\subsection{Formulation of the $SP_N$ Equations}
\label{subsec:spn_equations}
The $P_N$ equations give $N+1$ coupled first-order equations
capturing the spatial and angular-dependence of the solution. In
multiple dimensions, the equation set becomes large and coupled not
only through angular moments but also through the spatial
variables. As a simpler alternative to multidimensional $P_N$
solutions, Gelbard recognized in 1960 that the planar $P_N$ equations
could be simplified and applied an ad-hoc method to extend them to
multiple dimensions, yielding the $SP_N$ equations. These equations
are not only fewer in number, but also take on a diffusion-like form
while maintaining the angular character of the flux, making them
amenable to solutions with modern diffusion methods.

First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
equations by solving for the $n^th$ Legendre flux moment in the
odd-order equations:
\begin{equation}
  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
  \label{eq:odd_moments}
\end{equation}
for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
reduced group of equations for the even moments:
\begin{multline}
  -\frac{\partial}{\partial x}
  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
  0,2,4,\cdots,N\:.
  \label{eq:reduced_pn}
\end{multline}
Immediately, we note the diffusion-like nature of
Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
equations. To extend these equations to multiple dimensions, gelbard
simple replaced the x-plane spatial derivatives in the reduced set of
equations with general multidimensional gradient operators:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
  0,2,4,\cdots,N\:,
  \label{eq:spn_equations}
\end{multline}
yielding a multidimensional set of $(N+1)/1$ angular coupled equations
known as the $SP_N$. Again, we provide closure to this set of
equations with $\phi_{N+1} = 0$. As a concrete example, we will
consider the $SP_7$ equations:
\begin{subequations}
  \begin{gather}
    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
    \Sigma_0 \phi_0 = q \\ 
    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
    \Sigma_2 \phi_2 = 0\\
    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
    \Sigma_6 \phi_6 = 0 \:.
  \end{gather}
  \label{eq:sp7_equations}
\end{subequations}
To further modify these equations, we can use a change of variables to
create a new group of equations such that the gradients are operating
on a single vector:
\begin{subequations}
  \begin{gather}
    u_1 = \phi_0 + 2\phi_2 \\
    u_2 = 3\phi_2 + 4\phi_4 \\
    u_3 = 5\phi_4 + 6\phi_6 \\
    u_4 = 7\phi_6 \:,
  \end{gather}
  \label{eq:spn7_subs}
\end{subequations}
or equivalently:
\begin{subequations}
  \begin{gather}
    \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
    \frac{16}{35}u_4 \\
    \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
    \phi_4 = \frac{1}{5}u_3 - \frac{6}{35}u_4\\
    \phi_6 = \frac{1}{7}u_4\:.
  \end{gather}
  \label{eq:spn7_subs_inverse}
\end{subequations}
When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
\begin{subequations}
  \begin{gather}
    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
    = -q \\
    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
    \frac{1}{7}u_4 \Bigg] = 0 \:.
  \end{gather}
  \label{eq:spn7_subs_equations}
\end{subequations}
If we rearrange the Eq~(\ref{eq:spn7_subs_equations} such that only
one divergence operation is present in each equation, we can formulate
this as a matrix system of 4 equations in the case of the $SP_7$
approximation:
\begin{equation}
  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
  \label{eq:spn_matrix}
\end{equation}
with $\mathbf{u}$ the vector of solution variables:
\begin{equation}
  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
  \label{eq:spn7_solution_vector}
\end{equation}
$\mathbf{D}$ the vector of effective diffusion coefficients:
\begin{equation}
  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \

>>>>

%%---------------------------------------------------------------------------%%
\chapter{Derivation of the $P_N$ Equations}
\label{chap:pn_equations}
Here we derive the $P_N$ equations, a simplified form of the general
transport equation where Legendre polynomials are used to expand the
angular flux and scattering cross section variables as a means of
capturing the angular structure of the solution. Before deriving this
form of the transport equation, we briefly discuss a few properties of
Legendre polynomials that we will find useful in the derivation.

\section{Legendre Polynomials}
\label{sec:legendre_polys}
The Legendre polynomials are an orthogonal set of functions that
are solutions to Legendre's differential equation. They have the
following form \citep{lewis_computational_1993}:
\begin{equation}
  P_l(\mu) = \frac{1}{2^l l!}\frac{d^l}{d \mu^l}(\mu^2-1)^l\:.
  \label{eq:general_legendre_poly}
\end{equation}
These functions have several useful properties including
orthogonality:
\begin{equation}
  \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = \frac{1}{2l+1}\delta_{l l'}\:,
  \label{eq:legendre_orthog}
\end{equation}
a recurrence relation:
\begin{equation}
  \mu P_l(\mu) = \frac{1}{2l+1}[(l+1)P_{l+1}(\mu) + l P_{l-1}(\mu)]\:,
  \label{eq:legendre_recurrence}
\end{equation}
and an addition theorem:
\begin{equation}
  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}\sum_{m=-l}^l
  Y_{lm}(\hat{\Omega})Y^*_{lm}(\hat{\Omega}')\:,
  \label{eq:legendre_addition}
\end{equation}
where the functions $Y_{lm}(\hat{\Omega})$ are the spherical
harmonics. We can form the addition theorem in this way because the
spherical harmonics are in fact just harmonic multiples of the
Legendre polynomials:
\begin{equation}
  Y_{lm}(\hat{\Omega}) =
  \sqrt{\frac{(2l+1)(l-m)!}{(l+m)!}}P^m_l(\mu)e^{i \omega m}\:,
  \label{eq:spherical_harmonic}
\end{equation}
where $\omega$ is the azimuthal component of the streaming
direction. We can reduce Eq~(\ref{eq:legendre_addition}) for the
planar geometry we are studying by ignoring the azimuthal components
of the addition theorem. As shown in Eq~(\ref{eq:spherical_harmonic}),
the azimuthal dependence is given by the harmonic component, $e^{i
  \omega m}$, and therefore we choose to ignore all terms in
Eq.~(\ref{eq:legendre_addition}) where $m \neq 0$. This gives:
\begin{equation}
  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}
  Y_{l0}(\hat{\Omega})Y^*_{l0}(\hat{\Omega}')\:.
  \label{eq:legendre_addition_2}
\end{equation}
Per Eq~(\ref{eq:spherical_harmonic}) we have:
\begin{equation}
  Y_{l0}(\hat{\Omega}) = \sqrt{2l+1}P_l^0(\mu)\:,
  \label{eq:harmonic_0}
\end{equation}
where $P_l^0(\mu) = P_l(\mu)$ is the $0^{th}$ associated Legendre
function. Finally, we can reduce the addition theorem from
Eq~(\ref{eq:legendre_addition_2}) with Eq~(\ref{eq:harmonic_0}) to a
simple product for planar geometry:
\begin{equation}
  P_l(\hat{\Omega} \cdot \hat{\Omega}') = P_l(\mu)P_l(\mu')\:.
  \label{eq:legendre_addition_3}
\end{equation}

\section{Planar $P_N$ Equations}
\label{sec:planar_pn}
With the Legendre polynomial properties defined above, we can proceed
by deriving the $P_N$ equations for planar geometry. For these
equations, we will start by assuming a monoenergtic field of neutrons
such that we are solving the following reduced form of the transport
equation: 
\begin{equation}
  \mu \frac{\partial}{\partial x} \psi(x,\mu) + \sigma(x) \psi(x,\mu)
  = \int \sigma_s(x,\hat{\Omega}' \rightarrow \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}') d\Omega' + \frac{q(x)}{4 \pi}\:.
  \label{eq:mono_transport}
\end{equation}
The $P_N$ equations introduce the approximation that the angular
dependence of the scattering cross section, $\sigma_s$, and the angular
flux $\psi$, can be discretized by expanding them in Legendre
polynomials as follows:
\begin{equation}
  \psi(x,\mu) = \sum_{n=0}^\infty (2n+1)P_n(\mu)\phi_n(x)\:,
  \label{eq:flux_expansion}
\end{equation}
\begin{equation}
  \sigma_{sm}(x) = \sum_{m=0}^\infty (2m+1)P_m(\mu)\sigma_s(x)\:,
  \label{eq:scattering_expansion}
\end{equation}
where we have supressed the $2\pi$ generated by integrating away the
azimuthal angular component and $\phi_n(x)$ in
Eq~(\ref{eq:flux_expansion}) is referred to as the $n^{th}$ Legendre
moment of the neutron flux and is given by:
\begin{equation}
  \phi_n(x) = \int_{-1}^1 P_n(\mu)\psi(x,\mu)d\mu\:.
  \label{eq:legendre_moments}
\end{equation}
We first insert the expansions given by Eq~(\ref{eq:flux_expansion})
and (\ref{eq:scattering_expansion}) into the planar transport equation
given by Eq~(\ref{eq:mono_transport}):
\begin{multline}
  \frac{\partial}{\partial x}\Big[\sum_{n=0}^\infty (2n+1) \phi_n \mu
    P_n(\mu) \Big] + \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) =
  \\ \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d \mu' + q\:,
  \label{eq:pn_deriv_1}
\end{multline}
where the dependence on the spatial variable, $x$, has been
supressed. To arrive at the $P_N$ equations we multiply
Eq~(\ref{eq:pn_deriv_1}) by $P_m(\mu)$ and integrate over the angular
domain $\int_{-1}^1 d \mu$. We will look at each term in
Eq~(\ref{eq:pn_deriv_1}) individually.

@@

Now that we have expanded all angular dependent terms in
Eq~(\ref{eq:pn_deriv_1}) and reduced them appropriately, we can
combine them to generate the $P_N$ equations:
\begin{equation}
    \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{2n+1} \Big[
      (n+1) \phi_{n+1} + n \phi_{n-1} \Big] + \sum_{n=0}^\infty \sigma
    \phi_n = \sum_{n=0}^\infty \sigma_{sn} \phi_n + q\delta_{n0}\:.
  \label{eq:pn_deriv_18}
\end{equation}
More formally, the $P_N$ equations are written as:
\begin{equation}
   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
  \label{eq:final_pn_equations}
\end{equation}
where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
yields a set of $N+1$ equations for $N+2$ flux moments. We therefore
require an additional equation to close the system. In accordance with
the series truncation as an approximation we choose the last moment in
the expansion to be zero:
\begin{equation}
  \phi_{N+1} = 0\:.
  \label{eq:pn_closure}
\end{equation}
As an example, we will construct the P5 equations from
Eq~(\ref{eq:final_pn_equations}) and the closure given by
Eq~(\ref{eq:pn_closure}):
\begin{subequations}
  \begin{gather}
   \frac{\partial}{\partial x}\phi_{1} + \Sigma_0 \phi_0 = q\:,\\ 
   \frac{1}{3} \frac{\partial}{\partial x}\Big[ 2
     \phi_{2} + \phi_{0} \Big] + \Sigma_1 \phi_1 = 0\:,\\
   \frac{1}{5} \frac{\partial}{\partial x}\Big[ 3 \phi_{3} + 2
     \phi_{1} \Big] + \Sigma_2 \phi_2 = 0 \:,\\
   \frac{1}{7} \frac{\partial}{\partial x}\Big[ 4 \phi_{4} + 3
     \phi_{2} \Big] + \Sigma_3 \phi_3 = 0 \:,\\
   \frac{1}{9} \frac{\partial}{\partial x}\Big[ 5 \phi_{5} + 4
     \phi_{3} \Big] + \Sigma_4 \phi_4 = 0 \:,\\
   \frac{1}{11} \frac{\partial}{\partial x} 5 \phi_{4} + \Sigma_5
   \phi_5 = 0 \:.
  \end{gather}
  \label{eq:p5_equations}
\end{subequations}
This gives us a set of 6 coupled equations for the six Legendre
moments requested defined over the entire spatial domain for a single
energy group. In practice, only odd-numbered $P_N$ orders are
generally used \citep{lewis_computational_1993}. This is due to the
fact that using odd $N$ yields an even number of $N+1$ equations which
can be split evenly on the left and right boundaries of the problem to
facilitate the description of boundary conditions. We will choose this
convention when deriving the boundary conditions.

\section{Boundary Conditions for the $P_N$ Equations}
\label{subsbusec:bcs_pn}
Per the analysis in \citep{lewis_computational_1993}, two types of
boundary conditions will be discussed: reflecting and Marshak. Marshak
boundary conditions can be used to specify both vacuum conditions and
isotropic source conditions on the boundary.

@@
